---
title: "Project 1"
format: html
editor: visual
---

### Part 1: Data Processing 

#### Parsing Variable Values ###

Before analyzing any trends, we must first acquire and structure the raw data from the Census API. In this section, I develop a pipeline to fetch data, query the PUMS API, and format the results into a clean, multi-year dataset with assigned variable classes.

```{r}
# Load the required packages
library(httr)
library(jsonlite)

#function to fetch the JSON for a single year
get_json_dict <- function(year) {
  # Construct the specific URL for the given year
  url <- paste0("https://api.census.gov/data/", year, "/acs/acs1/pums/variables.json")
  
  #GET() to pull the data
  response <- GET(url)
  
  # Extract and parse from JSON into an R list
  parsed <- fromJSON(content(response, as = "text", encoding = "UTF-8"))
  
  # Return 'variables'
  return(parsed$variables)

}

#filter the list
filter_census <- function(full_list, year) {
  #core variables
  core_vars <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", 
                 "JWMNP", "PWGTP", "FER", "HHL", "SCH", 
                 "SCHL", "SEX", "REGION", "DIVISION")
  
  # The state variable
  if (year %in% c(2021, 2022)) {
    state_var <- "ST"
  } else {
    state_var <- "STATE"
  }
  
  # Combine core variables
  target_vars <- c(core_vars, state_var)
  
  #keep only the variables whose names match our target list
  filtered_list <- full_list[names(full_list) %in% target_vars]
  
  return(filtered_list)
}


#save the data for all required years
years <- c(2021, 2022, 2023, 2024)

# Create an empty list
census_dicts <- list()

# Loop through each year
for (y in years) {
  raw_list <- get_json_dict(y)
  
  filtered_list <- filter_census(raw_list, y)
  
  #Store List
  census_dicts[[as.character(y)]] <- filtered_list
}

# Save the final combined list as an .rds file 
saveRDS(census_dicts, file = "data/census.rds")

#readRDS("data/census.rds")

```

#### Parsing API Data


```{r}
my_api_key <- "c6758c38f49e633cc5c6631c8251876e5bc70c58"

example_url <- "https://api.census.gov/data/2024/acs/acs1/pums?get=SEX,PWGTP,MAR&for=state:*&SCHL=24&key=c6758c38f49e633cc5c6631c8251876e5bc70c58"

# make response object
response <- GET(example_url)
#response

#str(response)

raw_text <- content(response, as = "text")

parsed_census <- fromJSON(raw_text)
head(parsed_census)


```
#### Creating Helper Function

```{r}
library(dplyr)
library(tibble)

#Define function
format_census_tibble <- function(api_response) {
  raw_matrix <- fromJSON(content(api_response, as = "text"))
  col_names <- raw_matrix[1, ]
  data_rows <- raw_matrix[-1, ]
  clean_tibble <- as_tibble(data_rows)
  colnames(clean_tibble) <- col_names
  return(clean_tibble)
}

#Test on response
my_tibble <- format_census_tibble(response)
my_tibble

```


#### Wrapping in a function that parses data


```{r}
library(lubridate)
library(stringr)
library(purrr)

# Helper function to parse time strings and find midpoints
parse_time_midpoint <- function(time_string) {
  if (is.na(time_string) || time_string == "N/A" || str_detect(tolower(time_string), "not reported")) {
    return(NA) # Assign missing time values an NA[span_0](start_span)[span_0](end_span)
  }
  
  # Extract all times in the format "HH:MM a.m./p.m."
  times <- str_extract_all(tolower(time_string), "\\d{1,2}:\\d{2}\\s*[ap]\\.m\\.")[[1]]
  
  if (length(times) == 2) {
    # Convert to POSIXct for midpoint calculation
    t1 <- parse_date_time(times[1], "%I:%M %p")
    t2 <- parse_date_time(times[2], "%I:%M %p")
    
    # Calculate midpoint
    midpoint <- t1 + as.duration(t1 %--% t2) / 2
    return(format(midpoint, "%H:%M:%S")) # Return as standardized time string
  }
  return(NA)
}

# Main function to get and process census data
get_census_data <- function(year = 2024, 
                            num_vars = c("AGEP", "PWGTP"),
                            cat_vars = c("SEX"), 
                            geo_level = "state", 
                            geo_value = "37",
                            api_key = my_api_key) {
  
  # Year Validation
  if (!(year %in% c(2021, 2022, 2023, 2024))) {
    stop("Error: Year must be 2021, 2022, 2023, or 2024.")
  }
  
  # Validate Numeric
  valid_num_vars <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP")
  if (!("PWGTP" %in% num_vars)) {
    stop("Error: PWGTP must be included in numeric variables.")
  }
  if (length(num_vars) < 2) {
    stop("Error: You must select at least one numeric variable other than PWGTP.")
  }
  if (!all(num_vars %in% valid_num_vars)) {
    stop("Error: Invalid numeric variable selected.")
  }
  
  # Validate Categorical
  valid_cat_vars <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  if (length(cat_vars) < 1) {
    stop("Error: You must select at least one categorical variable.")
  }
  if (!all(cat_vars %in% valid_cat_vars)) {
    stop("Error: Invalid categorical variable selected.")
  }
  
  # Validate Geography
  if (!(geo_level %in% c("region", "division", "state")) || length(geo_level) > 1) {
    stop("Error: Geography level must be one of 'region', 'division', or 'state'.")
  }
  
  # Handle 'all'
  if (tolower(geo_value) == "all") {
    geo_value <- "*"
  }
  
  # Construct URL
  all_vars <- c(num_vars, cat_vars)
  var_string <- paste(all_vars, collapse = ",")
  base_url <- paste0("https://api.census.gov/data/", year, "/acs/acs1/pums")
  query_url <- paste0(base_url, "?get=", var_string, "&for=", geo_level, ":", geo_value, "&key=", api_key)
  
  # Request and format
  response <- GET(query_url)
  if (status_code(response) != 200) stop("Error: API request failed.")
  
  df <- format_census_tibble(response)
  
  #Load the dictionary
  census_dicts <- readRDS("data/census.rds")
  dict_year <- census_dicts[[as.character(year)]]
  
  # Process Categorical Variables
  for (cat_var in cat_vars) {
    if (cat_var %in% colnames(df)) {
      # Pull the appropriate year's variable level
      levels_info <- dict_year[[cat_var]]$values$item
      
      # Turn variables into factors
      df[[cat_var]] <- factor(df[[cat_var]], 
                              levels = names(levels_info), 
                              labels = unlist(levels_info))
    }
  }
  
  # 3. Process Numeric and Time Variables
  for (num_var in num_vars) {
    if (num_var %in% c("JWAP", "JWDP")) {
      # Match the API return codes to the dictionary string descriptions
      time_codes <- dict_year[[num_var]]$values$item
      mapped_strings <- time_codes[as.character(df[[num_var]])]
      
      # Turn variables into time values using the midpoint
      df[[num_var]] <- sapply(mapped_strings, parse_time_midpoint)
      df[[num_var]] <- hms::as_hms(df[[num_var]])
    } else {
      # Turn variables into numeric values
      df[[num_var]] <- as.numeric(df[[num_var]])
    }
  }
  
  # Add census class
  class(df) <- c("census", class(df))
  
  return(df)
}

#df
```
#### Combining Multiple Years

```{r}
#Function to pull years
get_multiple_years <- function(years = c(2023, 2024), 
                               num_vars = c("AGEP", "PWGTP"), 
                               cat_vars = c("SEX"), 
                               geo_level = "state", 
                               geo_value = "37",
                               api_key = my_api_key) {
  
  # Map over the requested years, call the single year function
  combined_df <- purrr::map_dfr(years, function(y) {
    single_year_data <- get_census_data(year = y, 
                                  num_vars =num_vars, 
                                  cat_vars = cat_vars,
                                  geo_level = geo_level,
                                  geo_value = geo_value,
                                  api_key = api_key)
    # Include a year variable
    single_year_data$YEAR <- as.factor(y) 
    return(single_year_data)
  })
  
  class(combined_df) <- c("census", class(combined_df))
  return(combined_df)
}

```

### Part 2: Writing a Generic Function for Summarizing

With the raw data cleanly formatted and assigned a custom census class, we can leverage R to handle our analysis efficiently. Below, I define custom summary() and plot() methods to calculate weighted statistics and generate weighted visualizations for our Census data.

```{r}
# Generic summary method for census objects
summary.census <- function(object, numeric_vars = NULL, categorical_vars = NULL, ...) {
  
  # Default to all numeric variables except PWGTP
  if (is.null(numeric_vars)) {
    numeric_vars <- names(object)[sapply(object, is.numeric)]
    numeric_vars <- setdiff(numeric_vars, "PWGTP")
  }
  
  # Default to all categorical variables
  if (is.null(categorical_vars)) {
    categorical_vars <- names(object)[sapply(object, is.factor)]
    if ("YEAR" %in% names(object)) {
      categorical_vars <- unique(c(categorical_vars, "YEAR"))
    }
  }
  
  results <- list()
  weights <- object$PWGTP
  total_weight <- sum(weights, na.rm = TRUE)
  
  # Numeric Summaries: weighted means and standard deviations
  for (var in numeric_vars) {
    if (!var %in% c("JWAP", "JWDP")) { 
      val <- object[[var]]
      
      weighted_mean <- sum(val * weights, na.rm = TRUE) / total_weight
      variance_term <- sum((val^2) * weights, na.rm = TRUE) / total_weight - (weighted_mean^2)
      weighted_sd <- sqrt(ifelse(variance_term > 0, variance_term, 0))
      
      results[[paste0(var, "_mean")]] <- weighted_mean
      results[[paste0(var, "_sd")]] <- weighted_sd
    }
  }
  
  # Categorical Summaries: weighted counts
  for (var in categorical_vars) {
    val <- object[[var]]
    counts <- tapply(weights, val, sum, na.rm = TRUE)
    results[[paste0(var, "_counts")]] <- counts
  }
  
  return(results)
}

#  plot method for census objects
library(ggplot2)

plot.census <- function(x, num_var, cat_var, ...) {
  # Boxplot of numeric variable by categorical variable using PWGTP weight
  p <- ggplot(x, aes(x = get(cat_var), y = get(num_var), weight = PWGTP)) + 
    geom_boxplot(fill = "steelblue", alpha = 0.7) +
    labs(x = cat_var, y = num_var, title = paste("Boxplot of", num_var, "by", cat_var)) +
    theme_minimal()
  
  return(p)
}
```
### Investigation

Investigating how monthly gas prices (GASP), a hot topic right now, and gross rent as a percentage of income (GRPIP) impact people in North Carolina. Adding to this, focusing on how these financial situations shift according to if someone is in school (SCH) and what level of education they have completed (SCHL).

To start, I will pull data for 2023 and 2024 for North Carolina.

```{r}

#pulling data in for criteria
data_nc <- get_multiple_years(years = c(2023,2024), num_vars = c("GASP", "GRPIP","PWGTP"), cat_vars = c("SCH", "SCHL"), geo_level = "state", geo_value = "37")


#Basic full summary
full_summary <- summary(data_nc)
full_summary
```
Testing my summary function 
```{r}
my_summary <- summary(data_nc, numeric_vars = c("GRPIP"), categorical_vars = c("SCH"))

head(my_summary$SCH)
```
Summary Statistics tell us basic stats and weights, plotting the distributions will give more understand to how these financial costs affect different groups. 

```{r}
# plotting Rent by School Enrollment
plot(data_nc, num_var = "GRPIP", cat_var = "SCH") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Distribution of NC Rent Burden by School Enrollment")


```
Now to look at Gas Prices vs Education level

```{r}
# Gas Costs by Education level
plot(data_nc, num_var = "GASP", cat_var = "SCHL") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Distribution of NC Monthly Gas Costs by Education Level")

```


