---
title: "Project 1"
format: html
editor: visual
---

### Introduction

Analyzing socioeconomic trends requires reliable, granular data. This project builds a custom R pipeline to query and process data from the Census Bureau's Public Use Microdata Sample (PUMS) API. Rather than downloading static datasets, we will develop functions that dynamically fetch, clean, and assign appropriate object classes to the data.
Ultimately, I will use this pipeline to investigate the financial pressures facing residents of North Carolina, specifically looking at how monthly gas costs and rent burdens vary across different levels of educational attainment and school enrollment.

### Part 1: Data Processing
 
#### Parsing Variable Values

Before querying the main database, we need a reliable way to interpret the API's numeric codes. I will first extract the variable dictionaries from the Census JSON endpoints for 2021 through 2024, filtering for our core demographic and economic variables.

```{r}
# Load the required packages
library(httr)
library(jsonlite)

#function to fetch the JSON for a single year
get_json_dict <- function(year) {
  # Construct the specific URL for the given year
  url <- paste0("https://api.census.gov/data/", year, "/acs/acs1/pums/variables.json")
  
  #GET() to pull the data
  response <- GET(url)
  
  # Extract and parse from JSON into an R list
  parsed <- fromJSON(content(response, as = "text", encoding = "UTF-8"))
  
  # Return 'variables'
  return(parsed$variables)

}

#filter the list
filter_census <- function(full_list, year) {
  #core variables
  core_vars <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", 
                 "JWMNP", "PWGTP", "FER", "HHL", "SCH", 
                 "SCHL", "SEX", "REGION", "DIVISION")
  
  # The state variable
  if (year %in% c(2021, 2022)) {
    state_var <- "ST"
  } else {
    state_var <- "STATE"
  }
  
  # Combine core variables
  target_vars <- c(core_vars, state_var)
  
  #keep only the variables whose names match our target list
  filtered_list <- full_list[names(full_list) %in% target_vars]
  
  return(filtered_list)
}


#save the data for all required years
years <- c(2021, 2022, 2023, 2024)

# Create an empty list
census_dicts <- list()

# Loop through each year
for (y in years) {
  raw_list <- get_json_dict(y)
  
  filtered_list <- filter_census(raw_list, y)
  
  #Store List
  census_dicts[[as.character(y)]] <- filtered_list
}

# Save the final combined list as an .rds file 
saveRDS(census_dicts, file = "data/census.rds")

#readRDS("data/census.rds")

```

#### Parsing API Data

```{r}
my_api_key <- "c6758c38f49e633cc5c6631c8251876e5bc70c58"

example_url <- "https://api.census.gov/data/2024/acs/acs1/pums?get=SEX,PWGTP,MAR&for=state:*&SCHL=24&key=c6758c38f49e633cc5c6631c8251876e5bc70c58"

# make response object
response <- GET(example_url)
#response

#str(response)

raw_text <- content(response, as = "text")

parsed_census <- fromJSON(raw_text)
head(parsed_census)


```

#### Creating Helper Function

```{r}
library(dplyr)
library(tibble)

#Define function
format_census_tibble <- function(api_response) {
  raw_matrix <- fromJSON(content(api_response, as = "text"))
  col_names <- raw_matrix[1, ]
  data_rows <- raw_matrix[-1, ]
  clean_tibble <- as_tibble(data_rows)
  colnames(clean_tibble) <- col_names
  return(clean_tibble)
}

#Test on response
my_tibble <- format_census_tibble(response)
my_tibble

```

#### Wrapping in a Function That Parses Data

With our dictionary saved, we can now build the main engine of our analysis. The get_census_data function below handles the API request, validates user inputs (such as ensuring multiple numeric variables are selected), and processes the raw matrix into a clean tibble. Crucially, it maps categorical variables to their factor levels using our saved dictionary and calculates the midpoint for time-based commute variables.
```{r}
library(lubridate)
library(stringr)
library(purrr)

# Helper function to parse time strings and find midpoints
parse_time_midpoint <- function(time_string) {
  if (is.na(time_string) || time_string == "N/A" || str_detect(tolower(time_string), "not reported")) {
    return(NA) # Assign missing time values an NA[span_0](start_span)[span_0](end_span)
  }
  
  # Extract all times in the format "HH:MM a.m./p.m."
  times <- str_extract_all(tolower(time_string), "\\d{1,2}:\\d{2}\\s*[ap]\\.m\\.")[[1]]
  
  if (length(times) == 2) {
    # Convert to POSIXct for midpoint calculation
    t1 <- parse_date_time(times[1], "%I:%M %p")
    t2 <- parse_date_time(times[2], "%I:%M %p")
    
    # Calculate midpoint
    midpoint <- t1 + as.duration(t1 %--% t2) / 2
    return(format(midpoint, "%H:%M:%S")) # Return as standardized time string
  }
  return(NA)
}

# Main function to get and process census data
get_census_data <- function(year = 2024, 
                            num_vars = c("AGEP", "PWGTP"),
                            cat_vars = c("SEX"), 
                            geo_level = "state", 
                            geo_value = NULL,
                            api_key = my_api_key) {
  
  # Year Validation
  if (!(year %in% c(2021, 2022, 2023, 2024))) {
    stop("Error: Year must be 2021, 2022, 2023, or 2024.")
  }
  
  # Validate Numeric
  valid_num_vars <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP")
  if (!("PWGTP" %in% num_vars)) {
    stop("Error: PWGTP must be included in numeric variables.")
  }
  if (length(num_vars) < 2) {
    stop("Error: You must select at least one numeric variable other than PWGTP.")
  }
  if (!all(num_vars %in% valid_num_vars)) {
    stop("Error: Invalid numeric variable selected.")
  }
  
  # Validate Categorical
  valid_cat_vars <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  if (length(cat_vars) < 1) {
    stop("Error: You must select at least one categorical variable.")
  }
  if (!all(cat_vars %in% valid_cat_vars)) {
    stop("Error: Invalid categorical variable selected.")
  }
  
  # Validate Geography
  if (!(geo_level %in% c("region", "division", "state")) || length(geo_level) > 1) {
    stop("Error: Geography level must be one of 'region', 'division', or 'state'.")
  }
  if (is.null(geo_value)){
    geo_value <- "37"
  }
  # Handle 'all'
  if (tolower(geo_value) == "all") {
    geo_value <- "*"
  }
  
  # Construct URL
  all_vars <- c(num_vars, cat_vars)
  var_string <- paste(all_vars, collapse = ",")
  base_url <- paste0("https://api.census.gov/data/", year, "/acs/acs1/pums")
  query_url <- paste0(base_url, "?get=", var_string, "&for=", geo_level, ":", geo_value, "&key=", api_key)
  
  # Request and format
  response <- GET(query_url)
  if (status_code(response) != 200) stop("Error: API request failed.")
  
  df <- format_census_tibble(response)
  
  #Load the dictionary
  census_dicts <- readRDS("data/census.rds")
  dict_year <- census_dicts[[as.character(year)]]
  
  # Process Categorical Variables
  for (cat_var in cat_vars) {
    if (cat_var %in% colnames(df)) {
      # Pull the appropriate year's variable level
      levels_info <- dict_year[[cat_var]]$values$item
      
      # Turn variables into factors
      df[[cat_var]] <- factor(df[[cat_var]], 
                              levels = names(levels_info), 
                              labels = unlist(levels_info))
    }
  }
  
  # 3. Process Numeric and Time Variables
  for (num_var in num_vars) {
    if (num_var %in% c("JWAP", "JWDP")) {
      # Match the API return codes to the dictionary string descriptions
      time_codes <- dict_year[[num_var]]$values$item
      mapped_strings <- time_codes[as.character(df[[num_var]])]
      
      # Turn variables into time values using the midpoint
      df[[num_var]] <- sapply(mapped_strings, parse_time_midpoint)
      df[[num_var]] <- hms::as_hms(df[[num_var]])
    } else {
      # Turn variables into numeric values
      df[[num_var]] <- as.numeric(df[[num_var]])
    }
  }
  
  # Add census class
  class(df) <- c("census", class(df))
  
  return(df)
}

#df
```

#### Combining Multiple Years

```{r}
#Function to pull years
get_multiple_years <- function(years = c(2023, 2024), 
                               num_vars = c("AGEP", "PWGTP"), 
                               cat_vars = c("SEX"), 
                               geo_level = "state", 
                               geo_value = "37",
                               api_key = my_api_key) {
  
  # Map over the requested years, call the single year function
  combined_df <- purrr::map_dfr(years, function(y) {
    single_year_data <- get_census_data(year = y, 
                                  num_vars =num_vars, 
                                  cat_vars = cat_vars,
                                  geo_level = geo_level,
                                  geo_value = geo_value,
                                  api_key = api_key)
    # Include a year variable
    single_year_data$YEAR <- as.factor(y) 
    return(single_year_data)
  })
  
  class(combined_df) <- c("census", class(combined_df))
  return(combined_df)
}

```

### Part 2: Writing a Generic Function for Summarizing

Our get_census_data function assigns a custom census class to the returned tibble. Because PUMS data uses the PWGTP variable to represent population weights (meaning one row represents multiple people), standard R summary functions will produce inaccurate statistical measures.

To solve this, I am using R's object-oriented system to build custom summary() and plot() methods. These generic functions will automatically detect our census objects and apply the correct weighted math to generate accurate means, standard deviations, and visualizations.

```{r}
# Generic summary method for census objects
summary.census <- function(object, numeric_vars = NULL, categorical_vars = NULL, ...) {
  
  # Default to all numeric variables except PWGTP
  if (is.null(numeric_vars)) {
    numeric_vars <- names(object)[sapply(object, is.numeric)]
    numeric_vars <- setdiff(numeric_vars, "PWGTP")
  }
  
  # Default to all categorical variables
  if (is.null(categorical_vars)) {
    categorical_vars <- names(object)[sapply(object, is.factor)]
    if ("YEAR" %in% names(object)) {
      categorical_vars <- unique(c(categorical_vars, "YEAR"))
    }
  }
  
  results <- list()
  weights <- object$PWGTP
  total_weight <- sum(weights, na.rm = TRUE)
  
  # Numeric Summaries: weighted means and standard deviations
  for (var in numeric_vars) {
    if (!var %in% c("JWAP", "JWDP")) { 
      val <- object[[var]]
      
      weighted_mean <- sum(val * weights, na.rm = TRUE) / total_weight
      variance_term <- sum((val^2) * weights, na.rm = TRUE) / total_weight - (weighted_mean^2)
      weighted_sd <- sqrt(ifelse(variance_term > 0, variance_term, 0))
      
      results[[paste0(var, "_mean")]] <- weighted_mean
      results[[paste0(var, "_sd")]] <- weighted_sd
    }
  }
  
  # Categorical Summaries: weighted counts
  for (var in categorical_vars) {
    val <- object[[var]]
    counts <- tapply(weights, val, sum, na.rm = TRUE)
    results[[paste0(var, "_counts")]] <- counts
  }
  
  return(results)
}

#  plot method for census objects
library(ggplot2)

plot.census <- function(x, num_var, cat_var, ...) {
  # Boxplot of numeric variable by categorical variable using PWGTP weight
  p <- ggplot(x, aes(x = get(cat_var), y = get(num_var), weight = PWGTP)) + 
    geom_boxplot(fill = "steelblue", alpha = 0.7) +
    labs(x = cat_var, y = num_var, title = paste("Boxplot of", num_var, "by", cat_var)) +
    theme_minimal()
  
  return(p)
}
```

### Investigation

Investigating how monthly gas prices (GASP), a hot topic right now, and gross rent as a percentage of income (GRPIP) impact people in North Carolina. Adding to this, focusing on how these financial situations shift according to if someone is in school (SCH) and what level of education they have completed (SCHL).

To start, I will pull data for 2023 and 2024 for North Carolina.

```{r}

#pulling data in for criteria
data_nc <- get_multiple_years(years = c(2023,2024), num_vars = c("GASP", "GRPIP","PWGTP"), cat_vars = c("SCH", "SCHL"), geo_level = "state", geo_value = "37")


#Basic full summary
full_summary <- summary(data_nc)
full_summary
```

Testing my summary function

```{r}
my_summary <- summary(data_nc, numeric_vars = c("GRPIP"), categorical_vars = c("SCH"))

head(my_summary$SCH)
```

Summary Statistics tell us basic stats and weights, plotting the distributions will give more understand to how these financial costs affect different groups.

```{r}
# plotting Rent by School Enrollment
plot(data_nc, num_var = "GRPIP", cat_var = "SCH") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Distribution of NC Rent Burden by School Enrollment")


```

#### Analyzing Rent Burden by School Enrollment

The distribution of Gross Rent as a Percentage of Income (GRPIP) reveals the housing pressure faced by different populations. Looking at the boxplot, we can see how the median rent burden and the spread of outliers compare between those currently enrolled in school and those who are not. We can see relatively similar spreads for all School Enrollment except Private school which is heavily skewed left with most having near 0 Rent Burden.

```{r}
# Gas Costs by Education level
plot(data_nc, num_var = "GASP", cat_var = "SCHL") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Distribution of NC Monthly Gas Costs by Education Level")

```

#### Analyzing Gas Costs by Education Level

Next, we map Monthly Gas Costs (GASP) against educational attainment (SCHL). Because commuting distances and vehicle types often correlate with employment types, which in turn correlate with education—we might expect to see variations in fuel spending. The visualization highlights that advanced degrees tend to have slightly higher Gas Costs compared to the rest.
